PANCHANGE_data_preprocessing

### 2020-08-17 PANCHANGE_data_preprocessing2020-07-15.Rmd
minor edits for RAMP data

#Packages
Install packages (if they are not available in your version of R)
```{r Installing packages}
#install.packages("summarytools")
#install.packages("tidyverse")
#install.packages("psych")
#install.packages("eeptools")
```

@@

## RAMP
```{r coping ramp ocir cols}
if(RAMP == TRUE) {
coping.ramp.ocir.raw <- readRDS(file = paste0(data.raw_path, "/ramp/ocir_coping_ramp.rds"))
dim(coping.ramp.ocir.raw)
colnames(coping.ramp.ocir.raw)
}
```

```{r coping ramp ocir summary labels}
if(RAMP == TRUE) {
summarytools::dfSummary(
  coping.ramp.ocir.raw,
  graph.col = F,
  valid.col = F
  #labels.col = F
)
}
```

@@

coping.ramp.ocir.raw.id <- coping.ramp.ocir.raw %>%

@@

    ocir.repeat_feel_numbers_retro = ocir.numbers_repeat_feel, ## Different from RAMP prepandemic
    ocir.clean_simply_wash_feel_retro = ocir.feel_contaminated_clean_simply, ## Different from RAMP prepandemic
    ocir.good_feel_bad_numbers_retro = ocir.bad_numbers_good_feel, ## Different from RAMP prepandemic
    ocir.hands_longer_wash_retro = ocir.longer_hands_wash, ## Different from RAMP prepandemic
    ocir.frequently_difficulty_rid_nasty_retro = ocir.frequently_rid_difficulty_nasty, ## Different from RAMP prepandemic
    ocir.strangers_touch_touched_difficult_retro = ocir.strangers_touch_touched_object, ## Different from RAMP prepandemic
    ocir.control_difficult_find_thoughts_retro = ocir.control_find_difficult_thoughts, ## Different from RAMP prepandemic
    ocir.i_collect_things_i_dont_need_retro = ocir.i_collect_things_i_dont_need., ## Different from RAMP prepandemic
    ocir.repeat_feel_numbers_base = ocir.numbers_repeat_feel.1, ## Different from RAMP prepandemic
    ocir.clean_simply_wash_feel_base = ocir.feel_contaminated_clean_simply.1, ## Different from RAMP prepandemic
    ocir.good_feel_bad_numbers_base = ocir.bad_numbers_good_feel.1, ## Different from RAMP prepandemic
    ocir.hands_longer_wash_base = ocir.longer_hands_wash.1, ## Different from RAMP prepandemic
    ocir.frequently_difficulty_rid_nasty_base = ocir.frequently_rid_difficulty_nasty.1, ## Different from RAMP prepandemic
    ocir.strangers_touch_touched_difficult_base = ocir.strangers_touch_touched_object.1, ## Different from RAMP prepandemic
    ocir.control_difficult_find_thoughts_base = ocir.control_find_difficult_thoughts.1, ## Different from RAMP prepandemic

@@

# Inspect dimensions
dim(coping.ramp.ocir.raw.id)
# Inspect colnames
colnames(coping.ramp.ocir.raw.id)
#Differences
dim(coping.ramp.ocir.raw)[1]-dim(coping.ramp.ocir.raw.id)[1]
}
```

```{r coping ramp ocir copy}
if(RAMP == TRUE) {
  coping.ocir.raw.id <- coping.ramp.ocir.raw.id
}
```

@@

Removing all prepandemic dataframes from NBR/RAMP (prepandemic measures don't exist for these cohorts)
```{r nbr/ramp data frame list}
if(NBR == TRUE | RAMP == TRUE) {
dfs.list <- list(
  dem.raw.id,
  mhd.raw.categories,
  coping.gad.raw.id,
  coping.pcl.raw.id,
  coping.phq.raw.id,
  coping.ocir.raw.id
)
}
```

@@

#Exclusion
Exclude those without sex, gender, age and incomplete PHQ/GAD items
Create a new new object for each exclusion step, e.g., data.no_sex, data.no_gender and data.no_sex.no_age
```{r Exclude participants that do not report sex}
dim(dat.raw)
dat.no.sex <- dat.raw %>%
  drop_na(Sex)
dim(dat.no.sex)

dim(dat.raw)[1]-dim(dat.no.sex)[1]
```
`r dim(dat.glad.raw)[1]-dim(dat.no.sex)[1]` were excluded because they did not report their sex

@@

summary(dat$phq.diff_score_prepan_base)
}
# Difference score for retrospective and baseline phq sum scores
dat <- dat %>%
  mutate(
    phq.diff_score_base_retro =
      as.numeric(
        phq.sum_score_base - phq.sum_score_retro  # baseline sum score MINUS retrospective sum score 
      )
  )

@@

summary(dat$gad.diff_score_prepan_base)
}
# Difference score for retrospective and baseline gad sum scores
dat <- dat %>%
  mutate(
    gad.diff_score_base_retro =
      as.numeric(
        gad.sum_score_base - gad.sum_score_retro  # baseline sum score MINUS retrospective sum score 
      )
  )

@@

```{r nbr/ramp data level groups PHQ}
if(NBR == TRUE | RAMP == TRUE){
dat <- dat %>%
  mutate(
    data_group_phq_numeric =
      case_when(
#        na_per_person_phq_base == 0 & na_per_person_phq_prepan == 0 & na_per_person_phq_retro == 0 ~ 5, #all complete baseline, prepan and retro
        na_per_person_phq_base != 0 & na_per_person_phq_retro == 0  ~ 4, #retro data only
        na_per_person_phq_base == 0 & na_per_person_phq_retro != 0  ~ 3, #baseline data only
#        na_per_person_phq_base == 0 & na_per_person_phq_prepan == 0 & na_per_person_phq_retro != 0  ~ 2, #prepan and baseline data only
        na_per_person_phq_base == 0 & na_per_person_phq_retro == 0 ~ 1,  #baseline and retro data only
#        na_per_person_phq_base != 0 & na_per_person_phq_prepan == 0 & na_per_person_phq_retro != 0  ~ 0  #prepan data only
      )
  )

@@

```{r nbr/ramp data level groups GAD}
if(NBR == TRUE | RAMP == TRUE){
dat <- dat %>%
  mutate(
    data_group_gad_numeric =
      case_when(
#        na_per_person_gad_base == 0 & na_per_person_gad_prepan == 0 & na_per_person_gad_retro == 0 ~ 5, #all complete baseline, prepan and retro
        na_per_person_gad_base != 0 & na_per_person_gad_retro == 0  ~ 4, #retro data only
        na_per_person_gad_base == 0 & na_per_person_gad_retro != 0  ~ 3, #baseline data only
#        na_per_person_gad_base == 0 & na_per_person_gad_prepan == 0 & na_per_person_gad_retro != 0  ~ 2, #prepan and baseline data only
        na_per_person_gad_base == 0 & na_per_person_gad_retro == 0 ~ 1,  #baseline and retro data only
#        na_per_person_gad_base != 0 & na_per_person_gad_prepan == 0 & na_per_person_gad_retro != 0  ~ 0  #prepan data only
      )
  )

@@

```{r nbr/ramp data level groups PCL}
if(NBR == TRUE | RAMP == TRUE){
dat <- dat %>%
  mutate(
    data_group_pcl_numeric =
      case_when(
#        na_per_person_pcl_base == 0 & na_per_person_pcl_prepan == 0 & na_per_person_pcl_retro == 0 ~ 5, #all complete baseline, prepan and retro
        na_per_person_pcl_base != 0 & na_per_person_pcl_retro == 0  ~ 4, #retro data only
        na_per_person_pcl_base == 0 & na_per_person_pcl_retro != 0  ~ 3, #baseline data only
#        na_per_person_pcl_base == 0 & na_per_person_pcl_prepan == 0 & na_per_person_pcl_retro != 0  ~ 2, #prepan and baseline data only
        na_per_person_pcl_base == 0 & na_per_person_pcl_retro == 0 ~ 1,  #baseline and retro data only
#        na_per_person_pcl_base != 0 & na_per_person_pcl_prepan == 0 & na_per_person_pcl_retro != 0  ~ 0  #prepan data only
      )
  )

@@

## RAMP respiratory
```{r ramp coping respiratory cols}
if(RAMP == TRUE){
ramp.respiratory.raw <- readRDS(file = paste0(data.raw_path, "/ramp/respiratory_coping_ramp.rds"))
dim(ramp.respiratory.raw)
colnames(ramp.respiratory.raw)
}
```

Select respiratory variables
```{r coping ramp respiratory cleaning}
if(RAMP == TRUE){
ramp.respiratory.raw.id <- ramp.respiratory.raw %>%
  drop_na(externalDataReference) %>% # Drop NAs
  distinct(externalDataReference, .keep_all = TRUE) %>% # Remove duplicates based on ID
  separate(externalDataReference, into = c("Sample", "ID"), sep = 4) %>% # Split ID in Sample and Number
  mutate(ID = as.numeric(ID)) %>%
  select(
    Sample, # Sample
    ID, # ID
    respiratory.breath_diarrhoea_fatigue_fever,
    respiratory.worst_ill_feel,
    respiratory.persistent_cough,
    respiratory.skipping_meals_because_you_felt_unwell,
    respiratory.was_your_sense_of_smell_impaired,
    respiratory.was_your_sense_of_taste_impaired,
    respiratory.feeling_tiredfatigued,
    respiratory.throat_swab_coronavirus_nose,
    respiratory.throat_swab_test_nose,
    respiratory.coronavirus_antibody_test,
    respiratory.results_antibody_test
  ) %>%
  add_numeric(., exclude = exclude_cols) %>%
  mutate_if(is.numeric, ~na_if(., -88)) %>% # Recode missing values to NAs in numeric variables
  mutate_if(is.numeric, ~na_if(., -99)) %>%
  mutate_if(is.numeric, ~na_if(., -77)) %>%
  mutate_if(is.factor, ~na_if(., "Seen but not answered")) %>% # Recode missing values to NAs in factor variables
  mutate_if(is.factor, ~na_if(., "Don't know")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to say")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Seen but not answered")) %>% # Drop empty factor levels
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Don't know")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to say"))

@@

++MD: Do we want to name the RAMP dataset something different, if the algorithm will need to be changed because they have categorical age and gender not sex? (otherwise can remove the if statement and rename the dataset 'dat.full' instead of dat.coping)
```{r clean respiratory dataset}
if(GLAD == TRUE | EDGI == TRUE | NBR == TRUE){
respiratory <- respiratory.raw.id

dat.coping <- left_join(
  x = dat,
  y = respiratory,
  by = "ID"
)
}
```

@@

++KT: Needs to be developed further for RAMP dataset
COVID prediction model for COPING
The COVID prediction model has coded the sexes opposite to our data set
```{r create numeric sex variable with male=1}
dat.coping <- dat.coping %>%
  mutate(
    Sex_respiratory =
      if_else(
        Sex == "Male",
        true = 1,
        false = 0
      )
  )
```

>>>>

#Packages
Install packages (if they are not available in your version of R)
```{r Installing packages}
#install.packages("summarytools")
#install.packages("tidyverse")
#install.packages("psych")
#install.packages("eeptools")
#install.packages("skimr")
```

@@

## RAMP
```{r coping ramp ocir cols}
if(RAMP == TRUE) {
ramp.ocir.raw <- readRDS(file = paste0(data.raw_path, "/ramp/ocir_ramp.rds"))
dim(ramp.ocir.raw)
colnames(ramp.ocir.raw)
}
```

```{r coping ramp ocir summary labels}
if(RAMP == TRUE) {
summarytools::dfSummary(
  ramp.ocir.raw,
  graph.col = F,
  valid.col = F
  #labels.col = F
)
}
```

@@

ramp.ocir.raw.id <- ramp.ocir.raw %>%
  drop_na(`Login ID`) %>% # Drop NAs
  distinct(`Login ID`, .keep_all = TRUE) %>% # Remove duplicates based on ID
  separate(`Login ID`, into = c("Sample", "ID"), sep = 4) %>% # Split ID in Sample and Number

@@

    ocir.felt_pandemic_feelings = ocir.feelings_pandemic_felt, # Comparison to before the pandemic
    ocir.saved_things_retro = ocir.saved_things.1, #reordered items to reflect questionnaire
    ocir.i_check_things_more_often_than_necessary_retro = ocir.i_check_things_more_often_than_necessary..1, #in RAMP .1 = retrospective
    ocir.arranged_properly_objects_upset_retro = ocir.arranged_properly_objects_upset.1,
    ocir.feel_compelled_count_things_retro = ocir.feel_compelled_count_things.1,
    ocir.strangers_touch_touched_difficult_retro = ocir.touch_touched_strangers_object.1,
    ocir.control_difficult_find_thoughts_retro = ocir.find_control_difficult_thoughts.1,
    ocir.i_collect_things_i_dont_need_retro = ocir.i_collect_things_i_dont_need..1,
    ocir.i_repeatedly_check_doors_windows_drawers_etc_retro = ocir.i_repeatedly_check_doors_windows_drawers_etc..1,
    ocir.arranged_things_change_upset_retro = ocir.arranged_things_change_upset.1,
    ocir.repeat_feel_numbers_retro = ocir.numbers_repeat_feel.1, 
    ocir.clean_simply_wash_feel_retro = ocir.feel_contaminated_clean_simply.1, 
    ocir.mind_upset_unpleasant_thoughts_retro = ocir.unpleasant_thoughts_mind_upset.1,
    ocir.afraid_avoid_throwing_things_retro = ocir.afraid_avoid_throwing_things.1,
    ocir.light_switches_water_taps_retro = ocir.repeatedly_check_gas_water.1,
    ocir.arranged_things_retro = ocir.arranged_things.1,
    ocir.good_feel_bad_numbers_retro = ocir.bad_numbers_good_feel.1, 
    ocir.hands_longer_wash_retro = ocir.longer_hands_wash.1, 
    ocir.frequently_difficulty_rid_nasty_retro = ocir.difficulty_rid_frequently_nasty.1, 
    ocir.saved_things_base = ocir.saved_things,
    ocir.i_check_things_more_often_than_necessary_base = ocir.i_check_things_more_often_than_necessary.,
    ocir.arranged_properly_objects_upset_base = ocir.arranged_properly_objects_upset,
    ocir.feel_compelled_count_things_base = ocir.feel_compelled_count_things,
    ocir.strangers_touch_touched_difficult_base = ocir.touch_touched_strangers_object, 
    ocir.control_difficult_find_thoughts_base = ocir.find_control_difficult_thoughts, 
    ocir.i_collect_things_i_dont_need_base = ocir.i_collect_things_i_dont_need.,
    ocir.i_repeatedly_check_doors_windows_drawers_etc_base = ocir.i_repeatedly_check_doors_windows_drawers_etc.,
    ocir.arranged_things_change_upset_base = ocir.arranged_things_change_upset,
    ocir.repeat_feel_numbers_base = ocir.numbers_repeat_feel, 
    ocir.clean_simply_wash_feel_base = ocir.feel_contaminated_clean_simply, 
    ocir.mind_upset_unpleasant_thoughts_base = ocir.unpleasant_thoughts_mind_upset,
    ocir.afraid_avoid_throwing_things_base = ocir.afraid_avoid_throwing_things,
    ocir.light_switches_water_taps_base = ocir.repeatedly_check_gas_water,
    ocir.arranged_things_base = ocir.arranged_things,
    ocir.good_feel_bad_numbers_base = ocir.bad_numbers_good_feel, 
    ocir.hands_longer_wash_base = ocir.longer_hands_wash, 
    ocir.frequently_difficulty_rid_nasty_base = ocir.difficulty_rid_frequently_nasty

@@

# Inspect dimensions
dim(ramp.ocir.raw.id)
# Inspect colnames
colnames(ramp.ocir.raw.id)
#Differences
dim(ramp.ocir.raw)[1]-dim(ramp.ocir.raw.id)[1]
}
```

```{r coping ramp ocir copy}
if(RAMP == TRUE) {
  coping.ocir.raw.id <- ramp.ocir.raw.id
}
```

@@

Removing all prepandemic dataframes from NBR/RAMP (prepandemic measures don't exist for these cohorts)
```{r nbr/ramp data frame list}
if(NBR == TRUE | RAMP == TRUE) {
dfs.list <- list(
  dem.raw.id,
  mhd.raw.categories,
  coping.gad.raw.id,
 # coping.pcl.raw.id, #WAITING FOR PCL DATA FOR RAMP
  coping.phq.raw.id,
  coping.ocir.raw.id
)
}
```

@@

#Exclusion
Exclude those without sex, gender, age and incomplete PHQ/GAD items
Create a new new object for each exclusion step, e.g., data.no_sex, data.no_gender and data.no_sex.no_age
```{r Exclude participants that do not report sex}
if(GLAD == TRUE | EDGI == TRUE | NBR == TRUE){
dim(dat.raw)
dat.no.sex <- dat.raw %>%
  drop_na(Sex)
dim(dat.no.sex)

dim(dat.raw)[1]-dim(dat.no.sex)[1]
}
```
`r dim(dat.glad.raw)[1]-dim(dat.no.sex)[1]` were excluded because they did not report their sex

```{r rename dataset for ramp}
if(RAMP == TRUE){
#there is no sex variable in ramp
dat.no.sex <- dat.raw
}
```

@@

summary(dat$phq.diff_score_prepan_base)
}

# Difference score for retrospective and baseline phq sum scores
dat <- dat %>%
  mutate(
    phq.diff_score_base_retro =
      as.numeric(
        phq.sum_score_base - phq.sum_score_retro  # baseline sum score MINUS retrospective sum score 
      )
  )

@@

summary(dat$gad.diff_score_prepan_base)
}

# Difference score for retrospective and baseline gad sum scores
dat <- dat %>%
  mutate(
    gad.diff_score_base_retro =
      as.numeric(
        gad.sum_score_base - gad.sum_score_retro  # baseline sum score MINUS retrospective sum score 
      )
  )

@@

```{r nbr/ramp data level groups PHQ}
if(NBR == TRUE | RAMP == TRUE){
dat <- dat %>%
  mutate(
    data_group_phq_numeric =
      case_when(
#        na_per_person_phq_base == 0 & na_per_person_phq_prepan == 0 & na_per_person_phq_retro == 0 ~ 5, #all complete baseline, prepan and retro
        na_per_person_phq_base != 0 & na_per_person_phq_retro == 0  ~ 4, #retro data only
        na_per_person_phq_base == 0 & na_per_person_phq_retro != 0  ~ 3, #baseline data only
#        na_per_person_phq_base == 0 & na_per_person_phq_prepan == 0 & na_per_person_phq_retro != 0  ~ 2, #prepan and baseline data only
        na_per_person_phq_base == 0 & na_per_person_phq_retro == 0 ~ 1  #baseline and retro data only
#        na_per_person_phq_base != 0 & na_per_person_phq_prepan == 0 & na_per_person_phq_retro != 0  ~ 0  #prepan data only
      )
  )

@@

```{r nbr/ramp data level groups GAD}
if(NBR == TRUE | RAMP == TRUE){
dat <- dat %>%
  mutate(
    data_group_gad_numeric =
      case_when(
#        na_per_person_gad_base == 0 & na_per_person_gad_prepan == 0 & na_per_person_gad_retro == 0 ~ 5, #all complete baseline, prepan and retro
        na_per_person_gad_base != 0 & na_per_person_gad_retro == 0  ~ 4, #retro data only
        na_per_person_gad_base == 0 & na_per_person_gad_retro != 0  ~ 3, #baseline data only
#        na_per_person_gad_base == 0 & na_per_person_gad_prepan == 0 & na_per_person_gad_retro != 0  ~ 2, #prepan and baseline data only
        na_per_person_gad_base == 0 & na_per_person_gad_retro == 0 ~ 1  #baseline and retro data only
#        na_per_person_gad_base != 0 & na_per_person_gad_prepan == 0 & na_per_person_gad_retro != 0  ~ 0  #prepan data only
      )
  )

@@

```{r nbr/ramp data level groups PCL}
if(NBR == TRUE | RAMP == TRUE){
dat <- dat %>%
  mutate(
    data_group_pcl_numeric =
      case_when(
#        na_per_person_pcl_base == 0 & na_per_person_pcl_prepan == 0 & na_per_person_pcl_retro == 0 ~ 5, #all complete baseline, prepan and retro
        na_per_person_pcl_base != 0 & na_per_person_pcl_retro == 0  ~ 4, #retro data only
        na_per_person_pcl_base == 0 & na_per_person_pcl_retro != 0  ~ 3, #baseline data only
#        na_per_person_pcl_base == 0 & na_per_person_pcl_prepan == 0 & na_per_person_pcl_retro != 0  ~ 2, #prepan and baseline data only
        na_per_person_pcl_base == 0 & na_per_person_pcl_retro == 0 ~ 1  #baseline and retro data only
#        na_per_person_pcl_base != 0 & na_per_person_pcl_prepan == 0 & na_per_person_pcl_retro != 0  ~ 0  #prepan data only
      )
  )

@@

## RAMP respiratory
```{r ramp coping respiratory cols}
if(RAMP == TRUE){
ramp.respiratory.raw <- readRDS(file = paste0(data.raw_path, "/ramp/respiratory_ramp.rds"))
dim(ramp.respiratory.raw)
colnames(ramp.respiratory.raw)
}
```

Select respiratory variables
```{r coping ramp respiratory cleaning}
if(RAMP == TRUE){
ramp.respiratory.raw.id <- ramp.respiratory.raw %>%
  drop_na(`Login ID`) %>% # Drop NAs
  distinct(`Login ID`, .keep_all = TRUE) %>% # Remove duplicates based on ID
  separate(`Login ID`, into = c("Sample", "ID"), sep = 4) %>% # Split ID in Sample and Number
  mutate(ID = as.numeric(ID)) %>%
  select(
    Sample, # Sample
    ID, # ID
    respiratory.breath_diarrhoea_fatigue_fever = respiratory.experienced_symptoms_unusual_shortness,
    respiratory.worst_ill_feel,
    respiratory.persistent_cough,
    respiratory.skipping_meals_because_you_felt_unwell,
    respiratory.was_your_sense_of_smell_impaired = respiratory.smell_impaired_sense_weeks,
    respiratory.was_your_sense_of_taste_impaired = respiratory.taste_impaired_sense_weeks,
    respiratory.feeling_tiredfatigued,
    respiratory.throat_swab_coronavirus_nose,
    respiratory.throat_swab_test_nose,
    respiratory.coronavirus_week_antibody_test,
    respiratory.results_antibody_test
  ) %>%
  add_numeric(., exclude = exclude_cols) %>%
  mutate_if(is.numeric, ~na_if(., -88)) %>% # Recode missing values to NAs in numeric variables
  mutate_if(is.numeric, ~na_if(., -99)) %>%
  mutate_if(is.numeric, ~na_if(., -77)) %>%
  mutate_if(is.factor, ~na_if(., "Seen but not answered")) %>% # Recode missing values to NAs in factor variables
  mutate_if(is.factor, ~na_if(., "Don't know")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to say")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Seen but not answered")) %>% # Drop empty factor levels
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Don't know")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to say"))

@@

++MD: Do we want to name the RAMP dataset something different, if the algorithm will need to be changed because they have categorical age and gender not sex? (otherwise can remove the if statement and rename the dataset 'dat.full' instead of dat.coping)
++KT: this is not the bit that needs changing for RAMP so have taken this out. 
```{r clean respiratory dataset}
respiratory <- respiratory.raw.id

dat.coping <- left_join(
  x = dat,
  y = respiratory,
  by = "ID"
)

@@

++KT: Needs to be developed further for RAMP dataset
COVID prediction model for COPING
The COVID prediction model has coded the sexes opposite to our data set
```{r create numeric sex variable with male=1}

dat.coping <- dat.coping %>%
  mutate(
    Sex_respiratory =
      if_else(
        Sex == "Male",
        true = 1,
        false = 0
      )
  )
```
